Informative Oceanographic Sampling using Excursion Probabilities for Multivariate Random Fields

### 2019-08-25 main_new.tex
Updates - synced Kanna and overleaf
Updates from Overleaf

% == Trygve Test
\usepackage{color, colortbl}
\definecolor{Gray}{gray}{0.9}
\newcommand{\kcomment}[1]{{\color{blue}{\{KR: #1\}}}}

@@

\begin{abstract}
  We examine experiment design for informative sampling for
  characterizing oceanographic phenomena. Methods are developed
  for efficient spatial sampling using autonomous robotic platforms,
  where the goal is to map the regions of cold and fresh water near a
  river mouth. Specifically, we define a design criterion based on
  reducing uncertainty in excursion sets of multivariate
  random fields. We develop analytical expressions for the expected
  variance reduction in excursion set indicators, and show how these can be
  applicable for static designs and sequential designs building on
  myopic and look-ahead planning. We use simulations to study the
  properties of the suggested methods. Results from field
  deployments with an autonomous underwater vehicle are also presented.

@@

Monitoring the worlds oceans has gained increased importance in the
light of the changing climate. While the focus has often been on
maga-fauna and the anthropogenic impact from micro-plastics to
pollution, biological oceanographers have focused intently on studying
micro-organisms at the base of the human food web. These organisms are
critically impacted by the changing dynamics in the upper water-column,
especially in coastal zones which are complex and often hard to observe
in space and time. By studying the bio-geochemical processes in the
upper water-column scientists can measure the impact of change, natural
or anthropomorphic, and provide an informed opinion to policy makers to
effect changes in preserving the environment. However, the challenge of 
% The pressure on marine resources is growing and increased accuracy,
% resolution, and persistent monitoring of the oceans is crucial for
% long-term sustainable management. A central problem for this type of
% environmental 
sensing in the small% in the large spatial extent
while studying the large spatio-temporal variation inherent to ocean
processes, prompts the need for effective means of sampling.
% allocation of sampling resources.
%\begin{em}
Sampling can, in this context, not simply be distributed evenly
---along simple transects, grids, or ``lawn-mover" patterns--- but must
instead be prioritized to relevant regions to ensure it is
cost-effective while providing adequate coverage of the area of
scientific interest.
% save resources and avoid
% redundant data.
%\end{em}

Traditional data collection at sea has typically been based on static
buoys, lagrangian floats or ship-centric methods, with significant
logistical limitations that directly impact coverage and sampling
resolution. Modern methods using satellite remote-sensing provide
large-scale coverage, but have limited resolution, are limited to the
sensing the surface of the ocean and impacted by cloud cover. With the
advent of robust mobile robotic platforms \cite{Bellingham07}, more
effective and higher resolution methods are making significant
contributions to environmental monitoring and sampling. 
% A
% sustained focus on prioritized and efficient data collection strategies
% have therefore started to emerge. The advent of marine robotic
% platforms, especially
In particular Autonomous underwater vehicles (AUVs), have made
significant strides in advancing the state of sampling
% provided means to execute this prioritization through the capacity of
% autonomy and data-driven sampling, where data collection in principle
% can be optimized. These capabilities have made
and consequently have made robotics an integral part of ocean
observation in large part driven by using 
% of the emerging sensing practice for ocean science, allowing scientists
% to increase the observational efficiency and resolution beyond what was
% previously possible. But how should a robotic platform, such as an AUV,
% effectively prioritize and identify important regions for sampling? The
% answer to this question relates to 
statistical methods in design and data assimilation
\cite{das11b,Graham2013,Das2015,das15,fossuminformation,fossum18b}.% , and
% the application domain is clearly an arena where statisticians can
% contribute.
% There has recently been some statistical attention in oceanography:
Others include \cite{wikle2013modern} who % present an overview of approaches for modeling
% uncertainties in the ocean sciences with
focus on hierarchical statistical models; \cite{sahu2008space} study
spatio-temporal models for sea surface temperature and salinity data,
while
% which will also are our key variables of interest in the current paper.
\cite{mellucci2018oceanic} look at the statistical prediction of
features using an underwater glider.

@@

The combination of statistical tools and robotic platforms is a natural
symbiosis which enables information-based sensing. Central to this is
the ability to model spatially correlated variables and provide formal
measures of uncertainty. By far the most practical models use Gaussian
processes (GPs), which will also be used in our formulation, in large
part because this approach allows efficient implementation and evaluation, enabling prioritization of
sampling efforts to be formalized into decision making systems on-board
a robotic platform.

@@

In this work, we study excursion sets (ES) and excursion probabilities
(EP) for spatial GPs, and the goal is to design sampling strategies for
reducing the variance in the ES indicators. The main contributions involve: i)
extending ES-based uncertainty criteria to multivariate random fields,
allowing the evaluation to include multiple environmental parameters
such as temperature and salinity, ii) develop methods for informative
environmental sensing and autonomous data collection, and iii)
demonstrating the approach in a real world scenario of sampling a river
plume using an AUV.

The remainder of the paper is organized as follows: Section \ref{sec:bg}
provides some background on ocean sampling, as also some notation for
ESs and EPs. Section \ref{sec:GP_EP} sets up the required assumptions of
GPs, including bi-variate random fields for temperature and salinity.
Section \ref{sec:sur} formulates the choice of sampling design criteria for
variance reduction in ES indicators, with analytical derivation for multivariate
spatial processes. Section \ref{sec:heuristics} builds on these sampling
criteria to present heuristics for adaptive sampling. Section
\ref{sec:simulations} shows results and discusses properties of the
methods in simulation while Section \ref{sec:case_study} instantiates
the work in the form of characterizing a river plume. We conclude with
Section \ref{sec:concl_disc} with a summary and a discussion of future
work.

@@

\subsection{Observing the Upper Ocean Column}

% The processes occurring in the upper water-column and the photic zone
% (upper sunlit zone) are of vital importance for long-term sustainable
% management of the greater marine ecosystem, climate, and human health.
% With the onset of the warming oceans and increased anthropogenic
% pressure, it is critical to improve our understanding of the associated
% ecosystem response.
Central to understanding the changes taking place in the upper
water-column is the understanding of biophysical interaction (see Fig.
\ref{fig:envir2}) driven by an agglomeration of physical forcing (wind,
topography, bathymetry, tidal influences and local geography such as
fresh water inflows) mixed with incipient micro-biology driven by
planktonic and coastal anthropogenic input, such as pollution and
agricultural runoff, which results in a range of ecosystem phenomena
(e.g. blooms, anoxic zones, plumes) ---with significant impact to
society. Renowned oceanographer Walter Munk called the
20\textsuperscript{th} century, the \emph{century of undersampling}
\citep{munk2002}, pointing out that lack of sufficient sampling
resolution in time and space, % owing to current observation practices.
% The lack of resolution, 
% arising from sparsely distributed sensing methods, is
as a critical issue for characterizing ocean phenomena. This view has
been driving ocean sensing practices towards more networked and
distributed use of robotic assets in order to fill the gaps in
our understanding (See Fig. \ref{fig:envir}).

\begin{figure}[!h] 
  \centering 
  \subfigure[Modern ocean sensing network, where robotic assets augment
  ship based measurements to increase coverage of the phenomena.]{\includegraphics[width =
    0.48\textwidth]{Figures/envir.pdf}\label{fig:envir1}}
  \hfill
  \subfigure[Different coastal processes directly or indirectly
  associated with gradients and boundaries.]{\includegraphics[width =
    0.48\textwidth]{Figures/envir_ocean.pdf}\label{fig:envir2}}
  \caption{Ocean observation is moving away from single-ship sampling
    towards more collaborative networked operations in order to resolve
    the numerous processes and their interaction.} \label{fig:envir}
\end{figure}

Our focus in this work is limited to spatial characterization of a river
plume system. Fig. \ref{fig:river_fronts} shows the survey area (the
outlet of the Nidelva river, Trondheim, Norway) taken as cold freshwater
enters from the river showing the separation of water masses and strong
gradients for both temperature and salinity. Because of the local
topography and the Coreolis force \citep{coriolis1835memoire} the cold
fresh water tends to stay near land to the east. Depending on the river
discharge, tidal effects, and temperature differences, this boundary in
water masses will get distorted. Prior knowledge about the location and
evolution of these features is therefore highly uncertain, making
deterministic planning sub-optimal.% , and one would rather demonstrate the
% ability of AUVs to map the plume area efficiently.

\begin{figure}[!h]
\centering
 % \subfigure[River front - Columbia River, Astoria, Oregon, US.]{\includegraphics[width = 0.45\textwidth]{Figures/pictures/a.png}\label{fig:river1}}
 % \hfill
 \subfigure[Tidal front - Korsfjorden, Norway.]{\includegraphics[width = 0.45\textwidth]{Figures/pictures/d.png}\label{fig:river2}}
 % \hfill
 % \subfigure[River plume - Rio de la Plata, Buenos Aires, Argentina.]{\includegraphics[width = 0.45\textwidth]{Figures/pictures/b.png}\label{fig:river3}}
 % \hfill
 \subfigure[Frontal patterns - Nidelva, Trondheim, Norway.]{\includegraphics[width = 0.45\textwidth]{Figures/pictures/c.png}\label{fig:river4}}
 \caption{Two examples of frontal features associated with
   river-ocean interaction. % (\ref{fig:river1}) The river plume front of
   % the Columbia River meeting the Pacific Ocean.
   (\ref{fig:river2})  Superimposed images (over 10 minutes) of a surface tidal front in
   Korsfjorden, Norway. % (\ref{fig:river3}) The river plume of the Rio de
   % la Plata, taken by European Space Agency (ESA)'s Envisat platform and
   % MEdium Resolution Imaging Spectrometer (MERIS) sensor.
   (\ref{fig:river4}) Aerial image of the mouth of the Nidelva river in
   Trondheim, Norway where fresh water meets the salty fjord and creates
   frontal patterns.} % Image courtesy: (\ref{fig:river1})
   % \cite{zamon2014marine}, (\ref{fig:river3}) ESA, CC BY-SA 3.0 IGO.}
 \label{fig:river_fronts}
\end{figure}

@@

River plumes belong to a class of ocean processes that are local and
usually act on smaller \emph{sub-mesoscale}, from 10 m -- 5 km where
spatial lateral variability tends to dominate. Remote sensing and
synthetic ocean models find it challenging to provide detailed at this
scale \citep{Lermusiaux:2006} and the principal way to resolve for fine
resolution is by direct observation. At larger \emph{mesoscale} ($>50$
km), both three-dimensional space and time dynamics are important and
can shift substantially, while in the case of river plumes one can often
limit scope to cover only lateral spatial elements ([north, east]
domain). Morover, for this case, time-effects can be regarded as static when
acquiring AUV data over a short time window and limited region. However,
the methodological framework can be extended to higher dimensional situations in a similar manner.

@@

AUVs operate underwater used to gather spatio-temporal data from a
targeted domain, carrying a scientific payload to survey the water
column or the seafloor. They run at $\sim 1-3$ m/s, can reach depths
from 100--6000m with in-water staying capacity depending on survey
speed, payload sensors and mission design. By deploying AUVs one can in
principle fill gaps and augment data obtained from either ocean models,
earth observing satellite data, or fixed-location sensors on buoys due
to their speed and mobility. Traditional operations and surveys with
AUVs are usually limited to planning samples along fixed transects or
survey lines set by human operators, programmed as a sequential plan
consisting of different waypoints and depths. But by using on-board
algorithms to continuously evaluate, update, and refine future sampling
locations, making information gathering adaptive \citep{das11b} is
feasible. For instance, an AUV could reconstruct or modify a survey line
based on what temperatures it is measuring by using embedded methods in
decision-theoretic planning \citep{Rajan12b}.

@@

% Determining paths for mobile robotic sensors in order to maximize the
% information gained about an environment is formally known as informative
% path planning. This, and related problems such as the orienteering
% problem \citep{Golden87}, has been studied in the context of graphs,
% where the potential measurement are assigned to nodes on which
% evaluation of different routes can be conducted.
%In this context, a fundamental result from \cite{nemhauser1978analysis} proves that a simple greedy algorithm (iteratively selecting the location which most increases the utility) can achieve a near-optimal solution if the problem can be shown to be \emph{submodular}\footnote{an intuitive diminishing returns property, where the informative value of adding sensors decrease with the number of sensors added.}. Building on this result, \cite{chekuri2005recursive} explored this in the setting of a graph using a recursive greedy algorithm, providing near-optimal solutions depending on the planning horizon and graph resolution. 
Adaptive sampling of an evolving frontal feature has been explored in
\cite{fronts11,Zhang2012,Pinto2018,costa19} using state-based autonomous
agents. These approaches typically use a reactive-adaptive scheme,
whereby exploration does not rely on a statistical model of the
environment, but rather adapts based on closing the sensing and
actuation loop. Myopic sampling, i.e. step-wise selection of the path
(on the graph) which reduces the information criterion the most, has
been used for adaptive surveys \citep{singh2009efficient,Binney2013}
which focus largely on reducing predictive variance or mutual
information (entropy). Because the variance and entropy reduction are
independent of the actual data realizations under the assumption of GP
models, the use of data-driven (adaptive) criteria was introduced to
include more targeted sampling of regions of scientific interest by
\cite{Low2009} and \cite{fossuminformation}.

@@

\subsection{Notation and excursion sets}

In our setting we study two variables; temperature and salinity, which
together characterize the phenomenon taking place in the river plume as
they define the gradients defining the frontal zone. At location
$\bx \in \mathcal{M} \subset \mathbf{R^2}$, we define the two process
variables by
$\bxi(\bx)=(\xi_a(\bx),\xi_b(\bx))$. %In practice there is of course time variability, but we assume this to be negligible during the AUV operation. Depth variability could also be important in some cases, but in the current application the with depths only a few meters this is considered irrelevant for mapping the phenomenon.

% In the context illustrated by Figure \ref{fig:river4},
Our focus is on spatial separation of cold freshwater from the saline
waters of the fjord and to be able to classify these water masses. In
this context, ES variability is a useful sampling criterion for guiding
the observational effort of the AUV. A bivariate ES is defined via joint
variables \kcomment{'being below thresholds' -- is this correct?}, at
individual locations, for the process over $\mathcal{M}$. If $T_a$ is
the threshold for temperature and $T_b$ the threshold for salinity,
we use the following definition:

 \begin{equation}\label{ES}
     \mbox{ES} = \{\bx \in \mathcal{M} | \xi_a(\bx) \leq T_a,\xi_b(\bx) \leq T_b\}.
 \end{equation}

 The excursions could also be defined as larger than a threshold, or
 within a boundary of limits for the two variables. Calculations and
 interpretations will be similar in these different cases.

The ES is defined by random indicator variables, and the associated EPs become 
\begin{equation}\label{eq:prob}
 p(\bx) = P(\xi_a(\bx) \leq T_a, \xi_b(\bx) \leq T_b), \hspace{3mm} \bx \in \mathcal{M}.
\end{equation}

The actual calculation of the EP would require a model
specification. As we demonstrate below, Gaussian modeling assumptions
facilitate the computation.  If the EPs in Eq. (\ref{eq:prob}) are
close to $1$ or $0$, it is easy to classify water masses; it is
challenging if the probabilities are close to $0.5$. In such a case it
would be worthwhile to gather data to improve classification skill.

Recent work in statistical literature on ESs and EPs connected to
spatial statistics includes
\citep{french2013spatio,bolin2015excursion,french2016credible}.  Most
are based on GPs for spatial phenomena, as in our work.  Our own focus
is on ES as defined in Eq. (\ref{ES}) and location-wise EPs and the
uncertainty reduction achieved by informative sampling.  In this sense
our work is similar to \cite{chevalier2014fast} and
\cite{azzimonti2016quantifying} who describe analytical results for
uncertainty reduction in ESs for univariate processes.  A major
contribution of our work is to derive closed-form results for
expectations of ESs in the situation with multivariate processes.

@@

@@

In our application with temperature and salinity, the solutions to the complex integral equations for the expected variance are reduced to calculating bi-variate and four-variate cumulative distribution functions for Gaussian vectors. For the more general situation one would be interested in $K$ random fields. Say, in the ocean sciences the variables could include chlorophyll or plankton variables, in addition to temperature and salinity.

>>>>

% == Trygve Test
\usepackage{color, colortbl}
\definecolor{Gray}{gray}{0.9}
\newcommand{\kcomment}[1]{{\color{blue}{\{KR: #1\}}}}
\newcommand{\tofcomment}[1]{{\color{green}{\{TF: #1\}}}}

@@

\begin{abstract}
Improving and optimizing oceanographic sampling is a crucial task for modern marine management and science faced with limited resources and highly dynamic processes. The combination of statistics and autonomous robotics provides opportunities for studying aspects related to experimental designs for effective characterization of oceanographic phenomena. In this paper methods are developed for efficient spatial sampling applied to the mapping of dynamic coastal processes by providing informative descriptions of spatial characteristics of ocean phenomena. Specifically, we define a design criterion based on reducing uncertainty in the excursion set of multivariate random fields, and derive analytical expressions for the expected Bernoulli variance reduction in the excursion set of spatial temperature and salinity variables. We show how these can be applicable in static designs and sequential designs, building on myopic and look-ahead sampling strategies. We use simulations to study the properties and differences between state-of-the-art approaches. Results from field deployments with an autonomous underwater vehicle are also presented, for a case study mapping the boundary of a river plume. The results demonstrate the ability of the method to effectively inform and execute data-driven environmental sampling.

@@

Monitoring the worlds oceans has gained increased importance in the light of the changing climate and increasing anthropogenic impact. A central problem for understanding these factors is the lack of representative data with sufficient resolution. Most of this \emph{undersampling} can be contributed to the large spatio-temporal variation on which ocean processes transpire, prompting the need for effective means of sampling. By \emph{sampling}, we refer to the design of observational strategies in the spatial domain, where the use of enabling autonomous platforms can be combined with statistical methods to pursuit measurements with high scientific relevance.

Traditional data collection at sea has typically been based on static buoys, lagrangian floats or ship-centric methods, with significant logistical limitations that directly impact coverage and sampling resolution. Modern methods using satellite remote-sensing provide large-scale coverage, but have limited resolution, are limited to the sensing the surface of the ocean, and is impacted by cloud cover. However, the advent of robust mobile robotic platforms \cite{Bellingham07} are making significant contributions to environmental monitoring and sampling. In particular autonomous underwater vehicles (AUVs), have made significant strides in advancing the state of sampling and consequently have made robotics an integral part of ocean observation in large part driven by using statistical methods in design and data assimilation \citep{das11b,Graham2013,Das2015,das15,fossuminformation,fossum18b}. Others include \cite{wikle2013modern} who focus on hierarchical statistical models; \cite{sahu2008space} study spatio-temporal models for sea surface temperature and salinity data, while \cite{mellucci2018oceanic} look at the statistical prediction of features using an underwater glider. Common to these approaches is the use of information-based methods to infer environmental sampling and a few important characteristics should be mentioned, namely:

\begin{itemize}
    \item Decisions on where to sample happen sequentially, meaning recent information can be assimilated and used to improve statistical models and reduce uncertainty associated with the  choices. 
    \item As a priori information about the state of the ocean is very uncertain, there lies a substantial efficiency gain in reacting to information obtained from measurements taken \emph{in-situ}, this is usually referred to as the \emph{adaptivity gap}.
    \item For this type of sampling problem, the number of choices (i.e. locations, paths, and candidate designs) makes the problem combinatorially large, creating a trade-off between optimization (finding the optimal design) and computability (arriving at a solution in reasonable time), which has to be considered in practice.
\end{itemize}{}

The combination of statistical tools and robotic platforms is a natural symbiosis which enables information-based sensing. Central to this is the ability to model spatially correlated variables and provide formal measures of uncertainty. By far the most practical models use Gaussian processes (GPs), which will also be used in our formulation, in large  part because this approach allows efficient implementation and evaluation, enabling prioritization of sampling efforts to be formalized into decision making systems on-board a robotic platform.

%Sampling can, in this context, not simply be distributed evenly ---along simple transects or ``lawn-mover" patterns--- but must instead be prioritized to relevant regions to ensure it is cost-effective while providing adequate coverage and resolution of the area of scientific interest.

%While the focus has often been on
%biological and anthropogenic impact from micro-plastics to
%pollution, biological oceanographers have focused intently on studying
%micro-organisms at the base of the human food web. These organisms are
%critically impacted by the changing dynamics in the upper water-column,
%especially in coastal zones which are complex and often hard to observe
%in space and time. By studying the bio-geochemical processes in the
%upper water-column scientists can measure the impact of change, natural
%or anthropomorphic, and provide an informed opinion to policy makers to
%effect changes in preserving the environment. However, the challenge of 
% The pressure on marine resources is growing and increased accuracy,
% resolution, and persistent monitoring of the oceans is crucial for
% long-term sustainable management.

% A
% sustained focus on prioritized and efficient data collection strategies
% have therefore started to emerge. The advent of marine robotic
% platforms, especially

% provided means to execute this prioritization through the capacity of
% autonomy and data-driven sampling, where data collection in principle
% can be optimized. These capabilities have made

% of the emerging sensing practice for ocean science, allowing scientists
% to increase the observational efficiency and resolution beyond what was
% previously possible. But how should a robotic platform, such as an AUV,
% effectively prioritize and identify important regions for sampling? The
% answer to this question relates to

% , and
% the application domain is clearly an arena where statisticians can
% contribute.
% There has recently been some statistical attention in oceanography:

@@

@@

In this work, we build on excursion sets (ES) and excursion probabilities (EP) for spatial design of experiments. The goal is to design sampling strategies for reducing the spatially integrated Bernoulli variance (IBV) in the ES indicators.  The main contributions involve: i) extending ES-based uncertainty criteria to multivariate GPs, allowing the evaluation to include multiple environmental parameters such as temperature and salinity, ii) developing methods for informative environmental sensing and autonomous data collection, and iii) demonstrating the approach in a real world scenario of sampling a river plume using an AUV.

The remainder of the paper is organized as follows: Section \ref{sec:bg} provides some background on ocean sampling. Section \ref{sec:ESEP} defines ESs and EPs, and the design criteria of IBV. Section \ref{sec:GP_EP} sets up the required statistical modeling assumptions of the bi-variate GPs for temperature and salinity. Section \ref{sec:sur} builds on these when deriving the analytical derivation of the design criteria. Section \ref{sec:heuristics} extends results to present heuristics for adaptive sampling. Section \ref{sec:simulations} shows results and discusses properties of the methods in simulation studies. Section \ref{sec:case_study} demonstrates the methodology in field work characterizing a river plume.  Section \ref{sec:concl_disc} contains summary and a discussion of future work.

@@

\subsection{Observing Processes in the Upper Ocean}

% The processes occurring in the upper water-column and the photic zone
% (upper sunlit zone) are of vital importance for long-term sustainable
% management of the greater marine ecosystem, climate, and human health.
% With the onset of the warming oceans and increased anthropogenic
% pressure, it is critical to improve our understanding of the associated
% ecosystem response.

% owing to current observation practices.

Central to understanding the changes taking place in the upper water-column is the understanding of biophysical interaction  driven by an agglomeration of physical forcing (wind, topography, bathymetry, tidal influences and local geography such as fresh water inflows) mixed with incipient micro-biology driven by planktonic and coastal anthropogenic input, such as pollution and agricultural runoff, which results in a range of ecosystem phenomena (e.g. blooms, anoxic zones, plumes) ---all influences with significant impact both directly and indirectly on society. Reverberating the requirement of more effective ocean sensing practices, renowned oceanographer Walter Munk referred to the 20\textsuperscript{th} century as the \emph{century of undersampling} \citep{munk2002}, pointing out that lack of sufficient sampling resolution in time and space, as a critical issue for characterizing ocean phenomena. Still today, decision makers struggle with limited observational efforts sometimes leading to severe environmental problems. The recent (May 2018) outbreak of harmful algal blooms in Northern Norway\footnote{Report from the Institute of Marine Research:https://bit.ly/2KZGJfI}, with losses exceeding 12 million tons of salmon valued at \$ 80 million (Norwegian Directorate of Fisheries), is a case in point in how adequate sampling efforts could have provided a more accurate warning assessment. These challenges of undersampling have been driving ocean sensing practices towards more networked and distributed use of robotic assets in order to fill the gaps in our understanding (See Fig. \ref{fig:envir}), where the surveying is guided by a model and the measurements made onboard the robot are used to update the model.

\begin{figure}[!h] 
  \centering 
  \subfigure[Modern ocean sensing network, where robotic assets augment ship based measurements to increase coverage of the phenomena.]{\includegraphics[width =
    0.48\textwidth]{Figures/envir.pdf}\label{fig:envir1}}
  \hfill
  \subfigure[Data-driven sampling under the sense-plan-act paradigm.]{\includegraphics[width =
    0.48\textwidth]{Figures/adaptive_cycle.pdf}\label{fig:sense-plan-act}}
  \caption{Ocean observation is moving away from single-ship sampling towards more autonomous and collaborative networked operations in order to resolve the numerous processes and their interaction.} \label{fig:envir}
\end{figure}

Focusing on coupling statistics and robotic sampling, our attention is in this work towards spatial characterization of a frontal systems, and more specifically fronts created by river plume systems. Fig. \ref{fig:nidelven} shows the survey area (the outlet of the Nidelva river, Trondheim, Norway) taken as cold freshwater enters from the river creating a strong separation gradient in both temperature and salinity. Because of the local topography and the Coreolis force \citep{coriolis1835memoire} the cold fresh water tends to flow near land to the east. Depending on the river discharge, tidal effects, wind and temperature differences, this boundary in water masses will get distorted. Prior knowledge about the location and evolution of these features is therefore highly uncertain, making deterministic planning sub-optimal.

\begin{figure}[!h] 
\centering 
\includegraphics[width=0.85\textwidth]{Figures/pictures/c-updated.pdf}
\caption{Frontal patterns - Nidelven, Trondheim, Norway.}
\label{fig:nidelven}
\end{figure}

%\begin{figure}[!h]
%\centering
 % \subfigure[River front - Columbia River, Astoria, Oregon, US.]{\includegraphics[width = 0.45\textwidth]{Figures/pictures/a.png}\label{fig:river1}}
 % \hfill
 %\subfigure[Tidal front - Korsfjorden, Norway.]{\includegraphics[width = 0.45\textwidth]{Figures/pictures/d.png}\label{fig:river2}}
 % \hfill
 % \subfigure[River plume - Rio de la Plata, Buenos Aires, Argentina.]{\includegraphics[width = 0.45\textwidth]{Figures/pictures/b.png}\label{fig:river3}}
 % \hfill
 %\subfigure[Frontal patterns - Nidelven, Trondheim, Norway.]{\includegraphics[width = 0.45\textwidth]{Figures/pictures/c.png}\label{fig:river4}}
 %\caption{Two examples of frontal features associated with
%   river-ocean interaction. % (\ref{fig:river1}) The river plume front of
   % the Columbia River meeting the Pacific Ocean.
%   (\ref{fig:river2})  Superimposed images (over 10 minutes) of a surface tidal front in
%   Korsfjorden, Norway. % (\ref{fig:river3}) The river plume of the Rio de
   % la Plata, taken by European Space Agency (ESA)'s Envisat platform and
   % MEdium Resolution Imaging Spectrometer (MERIS) sensor.
%   (\ref{fig:river4}) Aerial image of the mouth of the Nidelva river in
%   Trondheim, Norway where fresh water meets the salty fjord and creates
%   frontal patterns.} % Image courtesy: (\ref{fig:river1})
   % \cite{zamon2014marine}, (\ref{fig:river3}) ESA, CC BY-SA 3.0 IGO.}
% \label{fig:river_fronts}
%\end{figure}

@@

River plumes belong to a class of ocean processes that are local and usually act on smaller \emph{sub-mesoscale}, from 10 m -- 5 km where spatial lateral variability tends to dominate. Remote sensing and synthetic ocean models find it challenging to provide detailed at this scale \citep{Lermusiaux:2006} and the principal way to resolve for fine resolution is by direct observation. At larger \emph{mesoscale} ($>50$ km), both three-dimensional space and time dynamics are important and can shift substantially, while in the case of river plumes one can often limit scope to cover only lateral spatial elements ([north, east] domain). Moreover, for this case, time-effects can be regarded as static when acquiring AUV data over a short time window and limited region. However, the methodological framework can be extended to higher dimensional situations in a similar manner.

@@

AUVs operate underwater used to gather spatio-temporal data from a
targeted domain, carrying a scientific payload to survey the water
column or the seafloor. They run at $\sim 1-3$ m/s, can reach depths
from 100--6000m with in-water staying capacity depending on survey
speed, payload sensors and mission design. By deploying AUVs one can in
principle fill gaps and augment data obtained from either ocean models,
earth observing satellite data, or fixed-location sensors on buoys due
to their speed and mobility. Traditional operations and surveys with
AUVs are usually limited to planning samples along fixed transects or survey lines on a spatial grid, pre-programmed by the human operator to 
consist of different waypoints on the grid. But by using on-board
algorithms to continuously evaluate, update, and refine future sampling
locations, one makes information gathering adaptive \citep{das11b}.
The space of sampling opportunities is still limited to the spatial grid, but now there is much more flexibility in automatic adaption.
For instance, an AUV could reconstruct or modify a survey line
based on what temperatures it is measuring by using embedded methods in
decision-theoretic planning \citep{Rajan12b}.

@@

% Determining paths for mobile robotic sensors in order to maximize the
% information gained about an environment is formally known as informative
% path planning. This, and related problems such as the orienteering
% problem \citep{Golden87}, has been studied in the context of graphs,
% where the potential measurement are assigned to nodes on which
% evaluation of different routes can be conducted.
%In this context, a fundamental result from \cite{nemhauser1978analysis} proves that a simple greedy algorithm (iteratively selecting the location which most increases the utility) can achieve a near-optimal solution if the problem can be shown to be \emph{submodular}\footnote{an intuitive diminishing returns property, where the informative value of adding sensors decrease with the number of sensors added.}. Building on this result, \cite{chekuri2005recursive} explored this in the setting of a graph using a recursive greedy algorithm, providing near-optimal solutions depending on the planning horizon and graph resolution. 
Adaptive sampling of an evolving frontal feature has been explored in
\cite{fronts11,Zhang2012,Pinto2018,costa19} using state-based autonomous
agents. These approaches typically use a reactive-adaptive scheme,
whereby exploration does not rely on a statistical model of the
environment, but rather adapts based on closing the sensing and
actuation loop (Fig. \ref{fig:sense-plan-act}). Myopic sampling, i.e. step-wise selection of the path
(on the graph) which reduces the information criterion the most, has
been used for adaptive surveys \citep{singh2009efficient,Binney2013}
which focus largely on reducing predictive variance or mutual
information (entropy). Variance and entropy reduction are
independent of the actual data realizations under the assumption of GP models, and the use of data-driven (adaptive) criteria was introduced to include more targeted sampling of regions of scientific interest by
\cite{Low2009} and \cite{fossuminformation}.
In this paper, with the focus on mapping the river plume, designs that improve the classification of water masses should be rewarded.

@@

\section{Excursion sets and Integrated Bernoulli variance}
\label{sec:ESEP}

The objective is to characterize the river plume, focusing on spatial separation of cold freshwater from the warmer saline waters of the fjord. ESs and EPs are useful starting points to measure the ability to classify water masses, and the goal of experimental sampling is to improve this characterization, as quantified via changes in the EPs.

We consider a lateral domain, near sea surface, with locations $\bx \in \mathcal{M} \subset \mathbf{R^2}$. The two variables of interest, temperature and salinity, are defined as random processes; $\xi_a(\bx)$ denotes the temperature (in $^o C$) and $\xi_b(\bx)$ denotes the salinity (in $mg/l$) at location $\bx$.
We let $t_a$ be
a threshold for temperature and $t_b$ a threshold for salinity, the bivariate ES is then defined by
 \begin{equation}\label{ES}
     \mbox{ES} = \{\bx \in \mathcal{M} | \xi_a(\bx) \leq t_a,\xi_b(\bx) \leq t_b\}.
 \end{equation}
 The excursions could also be defined as larger than a threshold, or
 within a boundary of limits for the two variables. Calculations and
 interpretations will be similar in these different cases.

The ES holds random indicator variables, while the associated EPs are 
\begin{equation}\label{eq:prob}
 p(\bx) = P(\xi_a(\bx) \leq t_a, \xi_b(\bx) \leq t_b), \hspace{3mm} \bx \in \mathcal{M}.
\end{equation}
If the EPs in Eq. (\ref{eq:prob}) are
close to $1$ or $0$, it is easy to classify water masses; it is
challenging if the probabilities are close to $0.5$. 
The integrated Bernoulli variance (IBV), see \cite{bect2019}, is here defined as 
\begin{equation}\label{measVAR}
    V = \int_{\bx \in \mathcal{M}} p(\bx) \left(1-p(\bx)\right) d\bx.
\end{equation}
The measure is dominated by locations with EPs close to $0.5$.

If prior knowledge strongly indicates cold freshwater in a part of the domain, it is easy to classify, and there is likely little value in exploring that part much. Similarly with strong prior knowledge about warm saltwater. 
Data can be gathered at selected locations of the domain $\mathcal{M}$, to improve classification skill. We denote temperature measurements by $y_{a}(\bx_i)$ and salinity measurements by $y_{b}(\bx_i)$, at a sampling location $\bx_i \in \mathcal{M}$. Given a set of measurements, denoted $\by$, the conditional EPs are
\begin{equation}\label{eq:post_ep}
 p(\bx;\by) = P(\xi_a(\bx) \leq t_a, \xi_b(\bx) \leq t_b |\by). 
\end{equation}

Our goal is to construct designs for the data gathering, that are expected to improve the mapping of the water masses. 
When designing AUV sampling designs, it is hence natural to evaluate the expected reduction in the IBV, as provided by the survey data. In this way, we take the expectation with respect to the planned measurements $\by$;
\begin{equation}\label{sur}
    V_{\mbox{upd}} = \int_{\bx \in \mathcal{M}} E_{\by} \left\{ p(\bx;\by)\left( 1-p(\bx;\by)\right) \right\} d\bx, 
\end{equation}
where the probability is defined in Eq. (\ref{eq:post_ep}).
Considering a set $\mathcal{J}$ of possible designs for an AUV survey. A criterion for the selection is
\begin{equation}\label{crit}
    j^* = \mbox{argmin}_{j \in \mathcal{J}} V_{\mbox{upd}}(j),
\end{equation}
where the criterion $V_{\mbox{upd}}(j)$ in Eq. (\ref{sur}) is computed for each of the possible designs.

Recent works in statistical literature on ESs and EPs connected to
spatial statistics include
\cite{picheny2010,french2013spatio,bolin2015excursion,french2016credible}. Our focus
is on ES as defined in Eq. (\ref{ES}), and the
uncertainty reduction achieved by sampling as in Eq. (\ref{sur}).  In this sense
our work is similar to \cite{bect2012}, \cite{chevalier2014fast} and
\cite{azzimonti2016quantifying} who describe analytical results for
uncertainty reduction in ESs for univariate processes.  
A major contribution of our work is to derive closed-form results for this design criteria in the situation with multivariate GPs.

@@

The actual calculations of the EPs and the IBV require a model
specification. Here, Gaussian modeling assumptions are used, as described next. They allow closed form evaluation of the IBVs, as derived in Section \ref{sec:sur}.

@@

In our application with temperature and salinity, the solutions to the complex integral equations for the expected IBV are reduced to calculating bi-variate and four-variate cumulative distribution functions for Gaussian vectors. For the more general situation one would be interested in $K$ random fields. Say, in the ocean sciences the variables could include chlorophyll or other bio-chemical quantities, in addition to temperature and salinity.
